### 2017-07-22 chapters/introduction/patient.tex
Add latest. Remains to write about Fenics, biomarkers, and summary of papers

A general problem in the biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of and inflated version of the balloon. Having this picture in mind, it is
intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to obtain the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. We briefly
explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some
medical image, which is loaded with some loads $\mathbf{T}$. We want
to find the unloaded geometry 
$\Omega_{\mathcal{U}}$, so that when we load $\Omega_{\mathcal{U}}$ with $\mathbf{T}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{eq:intro_variational_form} with $\mathbf{T}$ representing
the traction on the endocardium, and which depends on the reference
coordinates. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To
find  $\Xvec_{\mathcal{U}}$ we let
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the load $\mathbf{T}$ do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge, especially if the tissue is soft and the pressure in
the right ventricle is large. In this case the
right ventricular free wall might collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{chapters/introduction/figures/unloading_fail/unloading_fail.png}
\caption{Backward displacement method might fail for bi-ventricular
  geometries. This can be seen be noticing that the RV free wall is
  buckling and colliding with the septum.}
\label{fig:unloading_fail}
\end{figure}

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides an a more stable
way of obtaining an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}.
This algorithm is based on subtracting a factor $k$ times
the displacement obtained after inflating the image-based geometry to
the target pressure. To estimate the
unloaded configuration $\Xvec_{\mathcal{U}}$, we perform the following steps:
\begin{enumerate}
  \item Inflate $\Xvec_{\mathcal{I}}$ to target
    pressure $(p_{\mathcal{I}}^{\mathrm{LV}},
    p_{\mathcal{I}}^{\mathrm{RV}})$ to obtain
    $\mathbf{d} = \mathbf{d}(\Xvec_{\mathcal{I}})$.
  \item Given $k \in \mathbb{R}$, the unloaded configuration is given by
    \begin{align}
      \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
     \end{align}
     Further let $\mathbf{d}_k = \mathbf{d}(\Xvec_{\mathcal{U}}(k))$
     be the displacement obtained after inflating the unloaded
     configuration to the target pressure.
   \end{enumerate}
The uloading algorithm is based on finding $k$ so that the difference
between the image-based geometry and the inflated, unloaded geometry
is minimized, that is 
\begin{align}
  \underset{k}{\text{minimize}}
  \;\;  \mathrm{res} ( \Xvec_{\mathcal{I}}, \Xvec_{\mathcal{U}}(k) + \mathbf{d}_k)
  \label{eq:unloading_Raghavan}
\end{align}
with
\begin{align}
  \mathrm{res}(A, B) = \sup_{a \in A} \inf_{b \in B} d(a, b), 
\end{align}
and $d$ being the Euclidean distance.

@@

\subsection{Data Assimilation}
The constitutive laws in cardiac mechanics comes from experiments
done on tissue slabs, and relates stress in the material to strain.
If we had the same data available for a given patient we could easily
have performed a regression analysis in order to identify the
parameters in the constitutive model. However, this type of data
typically requires that tissue samples are taken from the myocardium,
which is not an option when dealing with living humans. In general,
when dealing with physical systems such as the heart or weather
forecast, it is not always possible to design an experiment which
allows to determine all properties of a physical system
\cite{chapelle2013fundamental}. In such cases we have to take the
measurement that we have, and do our best to incorporate them into the
model.

@@

The other approach, called variational data assimilation, also known as
4D-var, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional. We will focus this approach since this is the approach
that is taken in this thesis. For a more thorough review of other
methods we refer to \cite{chapelle2013fundamental}.

@@

We will now explain the general setup and the main steps in variational data
assimilation. In order to keep consistent notation we will refer to
the \emph{state variable} as $\mathbf{w}$, which is our case
represents jointly the displacement and hydrostatic pressure
$\mathbf{w}=(\mathbf{u}, p)$.

The state variables are described by some \emph{physical forward
model} $\mathcal{M}$, which is our case is governed by the force balance
equation in \eqref{eq:force_balance_strong}. Moreover this model
typically depends on some parameters $\mu$ that we want to
adjust based on the observation at hand. For example, the unknown
parameters could be related to the stiffness of the myocardium which
which we want to adjust to each individual patient based on some
clinical observations. We can therefore assume that the underlying
model can be written in the form $\mathcal{M}(\mathbf{w}, \mu) = 0$.

We are also given a set of measurements (or observations) that we want
to assimilate, and we denote a single observation by $y$.
For example, $y$ might by a volume measurements at some given time
points in the cardiac cycle, measured using e.g ultrasound and image
processing techniques. The \emph{observation operator} $\mathcal{H}$,
is approximation of the observation and act as an operator from the
state space to the observation space. For a single observation, we
therefore have the relation
\begin{align}
  y = \mathcal{H}(\mathbf{w}) + \xi, 
\end{align}
where $\xi = \xi_{\mathbf{\mathcal{H}}} + \xi_{\mathbf{y}}$ represents
both the error in the measurements ($\xi_{\mathbf{y}}$) (e.g
noise in the data) and error in the representation of the observation
($\xi_{\mathbf{\mathcal{H}}}$) (that $\mathcal{H}$ do not represent
the data good enough).

@@

\begin{enumerate}
  \item For some initial guess $\mu_0$, solve the forward model
    $\mathcal{M} = 0$ to obtain $\state_0$.
  \item Compute $\mathcal{J}(\state_0, \mu_0)$.
  \item Compute the gradient $D \mathcal{J}(\state_0, \mu_0)$ to find
    the direction of the steepest descent.
  \item Move along the steepest direction and update the initial guess.
\end{enumerate}
The above steps are continued until we have reached a minimum for the
cost functional, e.g when the norm of the gradient becomes zero. In
practice we continue these steps until either the norm of the
gradient is below a certain tolerance or we have exceeded a maximum
number of iterations. Such a minimum may not always exists, or there
may be several local minima. In the latter case, the choice of initial
guess determines which local minima you will end up in. The data
assimilation pipeline is summarized in
Figure~\ref{fig:data_assimilation}.

\begin{remark}
  There exists gradient-free optimization methods that can be used to
  avoid computing the gradient. Examples include genetic algorithms or
  particle swarm algorithms. The drawback with such methods is that
  they typically requires a lot of functional evaluations. In our
  case this is very computationally expensive, because every functional
  evaluation require us to solve the non-linear forward model. 
\end{remark}

@@

\paragraph{The volume observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
In some cases, the ventricular cavity volume can easily be computed
analytically from the displacement field. Denote the inner cavity of
the left ventricle in the current configuration by
$\omega_{\mathrm{endo}}$. Further let $\mathrm{d} v$ and $\xvec$
denote an infinitesimal volume element and the coordinate in the
current configuration respectively. Then by the divergence theorem we have
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= \int_{\omega_{\mathrm{endo}}} \mathrm{d} v =
  \frac{1}{3}\int_{\omega_{\mathrm{endo}}} \nabla \cdot \xvec \mathrm{d} v =
  \frac{1}{3}\int_{\partial \omega_{\mathrm{endo}}} \xvec \cdot \mathbf{n} \mathrm{d} s, 
\end{align}
where $\mathbf{n}$ and $\mathrm{d} s$ are the unit normal and a surface
element on the current configuration respectively.
Note that the boundary $\partial \omega_{\mathrm{endo}}$ includes the
endocardial basal boundary. However in the case when the base is flat
and fixed in longitudinal direction, the contribution from this
boundary integral will be zero. In this case the boundary integral
will be the same as the integral over the endocardial boundary, with
the only change being the change of sign on the normal vector. By
Nansons formula we obtain 
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= -\frac{1}{3}\int_{\lvendo} \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S,
  \label{eq:volume_operator}
\end{align}
where now $\lvendo$,  $\Xvec$, $\Nvec$ and $\mathrm{d}S$ are respectively the
endocardial surface, the reference coordinate, the unit normal and a
surface element in the reference configuration.

@@

\begin{remark}
  Measured strains are computed relative to some reference
  geometry, which do not always coincide with the reference geometry
  chosen for your simulation (for example if you use the unloaded geometry). In
  these cases you have to either recompute the measured strain
  according to the chosen reference for the simulation, or recompute
  the simulated strains according to the correct reference geometry for
  your measurements. In the latter case, one compute the strain tensor
  using the modified deformation gradient $\tilde{\F} = \F
  \F_{\mathrm{ref}}^{-1}$ where $\F_{\mathrm{ref}}$ is the deformation
  gradient from your current reference to the correct reference state. 
\end{remark}

@@

\subsubsection{Computing the functional gradient} 
To move along the gradient descent it is required to actually
compute this functional gradient. The ``naive'' approach would be to
estimate the gradient using finite difference approximation. However,
this require one functional evaluation for each degree of freedom in
the parameter space, so when the number of parameters are large this
becomes impractical.

By the chain rule we have
\begin{align*}
  D  \widehat{\mathcal{J}}  =  D  \mathcal{J}(\state(\mu), \mu) =
   D_{\mu}\mathcal{J}(\state(\mu), \mu) + D_{\state}\mathcal{J}(\state(\mu), \mu) D_{\mu}\state(\mu).
\end{align*}
The terms $D_{\state}\mathcal{J}(\state(\mu), \mu)$ and
$D_{\mu}\mathcal{J}(\state(\mu), \mu)$ are typically straight forward
to compute. The term $D_{\mu}\state(\mu)$, on the other hand, is more
involved to compute.  If the take the total derivative of the
force balance equation, we obtain
\begin{align*}
  D \mathcal{M}(\state(\mu), \mu) =
  D_{\state} \mathcal{M}(\state(\mu), \mu) D_{\mu}\state(\mu)
  +  D_{\mu} \mathcal{M}(\state(\mu), \mu)  = 0,
\end{align*}
and we see that
\begin{align}
  D_{\state} \mathcal{M}(\state(\mu), \mu)  D_{\mu}\state(\mu) = - D_{\mu} \mathcal{M}(\state(\mu), \mu).
  \label{eq:tangent_linear_model}
\end{align}
The system \eqref{eq:tangent_linear_model} is known as the tangent
linear system, and can be computed numerically using ordinary
algorithmic differentiation techniques. However it requires that you
first specify the parameter $\mu$, meaning that it becomes expensive to
compute the functional gradient at several control points.

@@

Dolfin-Adjoint is a software package that based on the FEniCS project
which aims to derive the discrete adjoint equation and the tangent
linear models of finite element models implemented within the FEniCS
framework. While the traditional approach is to derive the adjoint code
from the forward code using automatic differentiation tools,
Dolfin-Adjoint utilizes the high level symbolic representation
\cite{UFL} in FEniCS to derive the discrete adjoint equations from
the discrete forward equations, and then the FEniCS system derives the
adjoint code from the discrete adjoint equations.

\subsubsection{Regularization}
Regularization is an important concept in many scientific disciplines
concerned with data fitting. When using data that essentially are
realizations of a stochastic process, we need to take into account
that the data are corrupted with noise. Moreover, the model we are using
is a simplification of a physical process, and fitting the
data exactly to the model is unrealistic. This is especially
important if the number of parameters we are trying to estimate are
more than the number of data points used as input. In such cases it is often
beneficial to make some assumptions about the solution we are looking
for in order to avoid overfitting. For example, we may assume that the
parameter we are searching for is smooth. Suppose that $\mu \in
H^1(\Omega)$ is the parameter we want to find, and that we want to
favor smooth solutions. Then the norm of the gradient $\| \nabla \mu
\|$ should be small. One way to achieve this is to add a penalty term
to the objective functional so that in stead of minimizing
$\mathcal{J}(\state, \mu)$ we minimize $\mathcal{J}(\state, \mu) +
\lambda \| \nabla \mu\|$, where the constant $\lambda$ controls the
amount of regularization and is called the regularization parameters.
Note that in the limit $\lambda \rightarrow \infty$ we would end up
with a constant function. This approach is referred to as Tikhonov
regularization.

@@

\subsection{Identifiability of parameters}
% Try dividing heart mesh is different segment, discuss uniqueness of
% paraperers
% Discuss regularization technices
When dealing with parameter estimation, you should always consider
questions about identifiability and uniqueness. This depends on the
model, the data and the objective functional
\cite{hadjicharalambous2015analysis}. A first test should be to
generate synthetic data with some prescribed parameters, feed this
error-free data to the data assimilation method, and ensure that you
are able to retrieve the parameters that generated that data. If you
end up with a different parameter set than the one generated the data,
your problem is not \emph{structurally identifiable}
\cite{chabiniok2016multiphysics}. Your problem is said to be
\emph{practical identifiable} if the parameters can be determined
uniquely based on the data at hand. If the test passes you should try
to add noise to the data, and see if your data assimilation is stable
with respect to noise in the measurement, and you should quantify the
model out error versus the model input error. Demonstrating identifiability is
often difficult, especially when your data is corrupted with noise,
and when the complexity of the model increases. Balancing the
complexity of model in terms of \emph{model fidelity}, i.e whether your
model is rich enough to representing the data, and the identifiability
is often key in order to have robust model. A too simple model will
often fail to represent your data, while a too complex model, might
give completely different answers with just a slight perturbation of
the data, i.e with noise added to the measurement. The same problem is
encountered in other scientific disciplines, for instance in
statistical learning theory, \emph{overfitting} is related to
overparameterization of the model, in which you can perfectly
represent your data, but a small perturbation will cause a significant
error. Another problem, which is also related to identifiability is
the question about uniqueness of the solution. A practical test is to
start the optimization algorithm from different initial points and see
if you end up at the same optimum. If this does not happen, it is
likely that the objective functional is not convex, which is a
requirement for uniqueness. Unfortunately, this is the case for many
problems in PDE-constrained optimization. In these cases, the objective
functional might have several distinct local minima, in which case
different techniques for find the best local minima exists
\cite{farrell2015multiple}.

\subsubsection{An example of a non-structurally identifiable problem}
Split the ventricle in two, and use volume as the only input.
Show that different parameter sets (with $\gamma$) will generate the
same volume

\subsubsection{Material parameter estimation}

\paragraph{One volume measurement only}

\paragraph{Multiple volume measurements}

>>>>

A general problem in the biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of and inflated version of the balloon. Having this picture in mind, it is
intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to estimate the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. We briefly
explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some
medical image, which is loaded with a pressure $p_{\mathcal{I}}^{\mathrm{LV}}$ (assume LV only of
simplicity). We want to find the unloaded geometry 
$\Omega_{\mathcal{U}}$, so that when we load $\Omega_{\mathcal{U}}$ with $p_{\mathcal{I}}^{\mathrm{LV}}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{eq:intro_variational_form} after applying the pressure
$p_{\mathcal{I}}^{\mathrm{LV}}$. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To find
$\Xvec_{\mathcal{U}}$ we let 
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the applied pressure, do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge, especially if the tissue is soft and the pressure in
the right ventricle is large. In this case the
right ventricular free wall might collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{chapters/introduction/figures/unloading_fail/unloading_fail.png}
\caption{Backward displacement method might fail for bi-ventricular
  geometries. This can be seen be noticing that the RV free wall is
  buckling and colliding with the septum. Here the image-based
  geometry is shown in black wire-frame, while the estimtated unloaded
geometry is shown in red.}
\label{fig:unloading_fail}
\end{figure}

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides a more stable
way of obtaining an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}.
This algorithm is based on subtracting a factor $k$ times
the displacement obtained after inflating the image-based geometry to
the target pressure. To estimate the
unloaded configuration $\Xvec_{\mathcal{U}}$, we perform the following steps:
\begin{enumerate}
  \item Inflate $\Xvec_{\mathcal{I}}$ to target
    pressure $(p_{\mathcal{I}}^{\mathrm{LV}},
    p_{\mathcal{I}}^{\mathrm{RV}})$ to obtain
    $\mathbf{d} = \mathbf{d}(\Xvec_{\mathcal{I}})$.
  \item Given $k \in \mathbb{R}$, the unloaded configuration is given by
    \begin{align}
      \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
     \end{align}
     Further let $\mathbf{d}_k = \mathbf{d}(\Xvec_{\mathcal{U}}(k))$
     be the displacement obtained after inflating the unloaded
     configuration to the target pressure.
   \end{enumerate}
The uloading algorithm is based on finding $k$ so that the difference
between the image-based geometry and the inflated, unloaded geometry
is minimized, that is 
\begin{align}
  \underset{k}{\text{minimize}}
  \;\;  \mathrm{res} ( \Xvec_{\mathcal{I}}, \Xvec_{\mathcal{U}}(k) + \mathbf{d}_k)
  \label{eq:unloading_Raghavan}
\end{align}
with the residual given by the direct Hausdorff distance \cite{huttenlocher1993comparing}
\begin{align}
  \mathrm{res}(A, B) = \sup_{a \in A} \inf_{b \in B} d(a, b), 
\end{align}
and $d$ being the Euclidean distance.

@@

\subsection{Data Assimilation}
The constitutive laws in cardiac mechanics comes from experiments
done on tissue slabs, and relates stress in the material to strain.
If we had the same data available for a given patient we could easily
have performed a regression analysis in order to identify the
parameters in the constitutive model. However, this type of data
typically requires that tissue samples are taken from the myocardium,
which is not an option when dealing with living humans. In general,
when dealing with physical systems such as the heart or the weather,
it is not always possible to design an experiment which
allows to determine all properties of a physical system
\cite{chapelle2013fundamental}. In such cases we have to take the
measurement that we have, and do our best to incorporate them into the
model.

@@

The other approach, called variational data assimilation, also known as
4D-var, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional. We will focus this approach since this is the approach
taken in this thesis. For a more thorough review of other
data assimilation techniques we refer to \cite{chapelle2013fundamental}.

@@

We will now explain the general setup and the main steps in variational data
assimilation. We will refer to
the \emph{state variable} as $\mathbf{w}$, which is our case
represents jointly the displacement and hydrostatic pressure
$\mathbf{w}=(\mathbf{u}, p)$.

The state variables are described by some \emph{physical forward
model} $\mathcal{M}$, which is our case is governed by the force balance
equation in \eqref{eq:force_balance_strong}. Moreover this model
typically depends on some parameters $\mu$ that we want to
adjust based on the observation at hand. We will refer to this
parameter as the \emph{control parameters}. For example, the control
parameters could be related to the stiffness of the myocardium which
we want to adjust to each individual patient based on some
clinical observations. We can therefore assume that the underlying
model can be written in the form $\mathcal{M}(\mathbf{w}, \mu) =
0$. Note that the state variable $\mathbf{w}$ typically depends on the
control parameters $\mu$, but we do not write this explicitly to
ease notation.

We are also given a set of measurements (or observations) that we want
to assimilate, and we denote a single observation by $y$.
For example, $y$ might by a volume measurements at some given time
point in the cardiac cycle, measured using e.g ultrasound and image
processing techniques. The \emph{observation operator} $\mathcal{H}$,
is approximation of the observation, and act as an operator from the
state space to the observation space. For a single observation, we
therefore have the relation
\begin{align}
  y = \mathcal{H}(\mathbf{w}) + \xi, 
\end{align}
where $\xi = \xi_{\mathbf{\mathcal{H}}} + \xi_{\mathbf{y}}$ represents
both the error in the measurements ($\xi_{\mathbf{y}}$) (e.g
noise in the data) and error in the representation of the observation
($\xi_{\mathbf{\mathcal{H}}}$) (that $\mathcal{H}$ do not represent
the data good enough).

@@

\begin{enumerate}
  \item For some initial guess $\mu_0$, solve the forward model
    $\mathcal{M} = 0$ to obtain $\state_0$.
  \item Compute $\mathcal{J}(\state_0, \mu_0)$.
  \item Compute the gradient $D \mathcal{J}(\state_0, \mu_0)$ to find
    the direction of the steepest descent.
  \item Move along the steepest direction and update the initial
    guess.
\end{enumerate}
The above steps are continued until we have reached a minimum for the
cost functional, e.g when the norm of the gradient becomes zero. In
practice we continue these steps until either the norm of the
gradient is below a prescribed tolerance, or the number of iterations
exceed a maximum number. Using this gradient descent approach will
always bring you towards a minimum. However, the minimum you find
might not be a global minimum. In this case, the choice of initial
guess determines which local minima you will end up in. If you can
prove that your cost functional is convex with respect to the control
parameters, you known that the minimum you find is a global
one.

The data assimilation pipeline is summarized in
Figure~\ref{fig:data_assimilation}.

\begin{remark}
  There exists gradient-free optimization methods that can be used to
  avoid computing the gradient. Examples include genetic algorithms or
  particle swarm algorithms. The drawback with such methods is that
  they typically requires a lot of functional evaluations. In our
  case this is very computationally expensive, because every functional
  evaluation requires a solution of the non-linear forward model. 
\end{remark}

@@

\paragraph{The volume observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
In some cases, the ventricular cavity volume can easily be computed
analytically from the displacement field $\uvec$. Denote the inner
cavity of the left ventricle in the current configuration by
$\omega_{\mathrm{endo}}$. Further let $\mathrm{d} v$ and $\xvec$
denote an infinitesimal volume element and the coordinate in the
current configuration respectively. Then by the divergence theorem we have
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= \int_{\omega_{\mathrm{endo}}} \mathrm{d} v =
  \frac{1}{3}\int_{\omega_{\mathrm{endo}}} \nabla \cdot \xvec \mathrm{d} v =
  \frac{1}{3}\int_{\partial \omega_{\mathrm{endo}}} \xvec \cdot \mathbf{n} \mathrm{d} s, 
\end{align}
where $\mathbf{n}$ and $\mathrm{d} s$ are the unit normal and a surface
element on the current configuration respectively.
Note that the boundary $\partial \omega_{\mathrm{endo}}$ includes the
endocardial basal boundary. However in the case when the base is flat
and fixed in longitudinal direction, the contribution from this
boundary integral will be zero. In this case the boundary integral
will be the same as the integral over the endocardial boundary, with
the only change being the change of sign on the normal vector. By
Nansons formula we obtain 
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= -\frac{1}{3}\int_{\lvendo} \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S,
  \label{eq:volume_operator}
\end{align}
where now $\lvendo$,  $\Xvec$, $\Nvec$ and $\mathrm{d}S$ are respectively the
endocardial surface, the reference coordinate, the unit normal and a
surface element in the reference configuration.

@@

\begin{remark}
  Measured strains are computed relative to some reference
  geometry, which do not always coincide with the reference geometry
  chosen for your simulation (for example if you use the unloaded geometry). In
  these cases you have to either recompute the measured strain
  according to the chosen reference for the simulation, or recompute
  the simulated strains according to the correct reference geometry for
  your measurements. In the latter case, the strain tensor can be
  computed using the modified deformation gradient $\tilde{\F} = \F
  \F_{\mathrm{ref}}^{-1}$ where $\F_{\mathrm{ref}}$ is the deformation
  gradient from your current reference to the correct reference state. 
\end{remark}

@@

\subsubsection{Computing the functional gradient} 
To move along the gradient descent it is required to actually
compute this functional gradient. The ``naive'' approach would be to
estimate the gradient using finite difference approximation. However,
this requires one functional evaluation for each degree of freedom in
the parameter space, so when the number of parameters are large this
becomes impractical.

By the chain rule we have
\begin{align*}
  D  \widehat{\mathcal{J}}  =  D  \mathcal{J}(\state(\mu), \mu) =
   D_{\mu}\mathcal{J}(\state(\mu), \mu) + D_{\state}\mathcal{J}(\state(\mu), \mu) D_{\mu}\state(\mu).
\end{align*}
The terms $D_{\state}\mathcal{J}(\state(\mu), \mu)$ and
$D_{\mu}\mathcal{J}(\state(\mu), \mu)$ are typically straight forward
to compute. The term $D_{\mu}\state(\mu)$, on the other hand, is more
involved to compute.  If we take the total derivative of the
force balance equation, we obtain
\begin{align*}
  D \mathcal{M}(\state(\mu), \mu) =
  D_{\state} \mathcal{M}(\state(\mu), \mu) D_{\mu}\state(\mu)
  +  D_{\mu} \mathcal{M}(\state(\mu), \mu)  = 0,
\end{align*}
and we see that
\begin{align}
  D_{\state} \mathcal{M}(\state(\mu), \mu)  D_{\mu}\state(\mu) = - D_{\mu} \mathcal{M}(\state(\mu), \mu).
  \label{eq:tangent_linear_model}
\end{align}
The system \eqref{eq:tangent_linear_model} is known as the tangent
linear system, and can be computed numerically using ordinary
algorithmic differentiation techniques. However it requires that you
first specify the parameter $\mu$, meaning that it becomes expensive to
compute the functional gradient at several control points.

@@

Dolfin-Adjoint \cite{farrell2013automated} is a software package that
based on the FEniCS project 
which aims to derive the discrete adjoint equation and the tangent
linear models of finite element models implemented within the FEniCS
framework. While the traditional approach is to derive the adjoint code
from the forward code using automatic differentiation tools,
Dolfin-Adjoint utilizes the high level symbolic representation
\cite{alnaes2014unified} in FEniCS to derive the discrete adjoint
equations from the discrete forward equations, and then the FEniCS
system derives the adjoint code from the discrete adjoint equations. 
For more information about Dolfin-Adjoint the reader is referred to
the official web page (\url{http://www.dolfin-adjoint.org/en/latest/}).

\subsubsection{Multiobjetive optimization}
Parameters in the model are estimated based on minimizing a cost
functional representing the mismatch between observations and
simulations. Assume we are given $N$ different observations, then in
principle we have $N$ different cost functionals to minimize:
\begin{equation}
  \begin{aligned}
    \label{eq:multiobjetive_opt}
    & \underset{\mu \in P}{\text{minimize}}
    & &  \left\{ \mathcal{J}_1(\state, \mu), \mathcal{J}_2(\state, \mu), \cdots, \mathcal{J}_N(\state, \mu) \right\} \\
    & \text{subject to}
    & &\mathcal{M}(\state, \lambda) = 0
  \end{aligned}
\end{equation}
with
\begin{align}
  \mathcal{J}_i(\mu) = \left( \mathbf{y} - \mathcal{H}(\state(\lambda)) \right)^2, \; i = 1,\cdots, N
\end{align}
The above observations comes from the same source, and hence it should
be intuitive that minimizing one of them, would also bring the other
closer to a minimum. However, with the presence of noise in the data,
this might not always be the case, and hence we have conflicting
objectives. Problem \eqref{eq:multiobjetive_opt} is called a
multiobjetive optimization problem \cite{deb2016multi}. Such problems
typically do not have a single optimal solution, but a family of
solutions called \emph{Pareto optimal solutions}. The basic methods for
solving such problems includes \emph{the weighted method} and \emph{the
  $\varepsilon$-constraint method}. In this thesis we have only used
the weighted method, which is based on minimizing a weighted sum of
the individual cost functionals, i.e:

\begin{equation}
  \begin{aligned}
    \label{eq:weighted_method}
    & \underset{\mu \in P}{\text{minimize}}
    & &  \sum_{i=1}^N \lambda_i \mathcal{J}_i(\state, \mu) \\
    & \text{subject to}
    & &\mathcal{M}(\state, \mu) = 0.
  \end{aligned}
\end{equation}
Here $\lambda_i$ represents a weighting of the different objectives,
and the choice of $\lambda_i$ should be based on 1) the importance of
objective $i$ and 2) the relative size of the cost functional
$\mathcal{J}_i$. It is also common to require $\sum_i \lambda_i = 1$
so that the weighted sum is a convex combination of the objective
functionals.

\subsubsection{Regularization}
Regularization is an important concept in many scientific disciplines
concerned with data fitting. When using data that essentially are
realizations of a stochastic process, we need to take into account
that the data are corrupted with noise. Moreover, the model we are using
is a simplification of a physical process, and fitting the
data exactly to the model is unrealistic. This is especially
important if the number of parameters we are trying to estimate are
more than the number of data points used as input. In such cases it is often
beneficial to make some assumptions about the solution we are looking
for in order to avoid overfitting. For example, we may assume that the
parameter we are searching for is smooth. Suppose that $\mu \in
H^1(\Omega)$ is the parameter we want to find, and that we want to
favor smooth solutions. Then the norm of the gradient $\| \nabla \mu
\|$ should be small. One way to achieve this is to add a penalty term
to the objective functional so that in stead of minimizing
$\mathcal{J}(\state, \mu)$ we minimize $\mathcal{J}(\state, \mu) +
\lambda \| \nabla \mu\|$, where the constant $\lambda$, known as the
regularization parameter, controls the amount of regularization.
Note that in the limit $\lambda \rightarrow \infty$ we would end up
with a constant function. This approach is referred to as Tikhonov
regularization.

@@

\subsection{Identifiability of parameters}
% Try dividing heart mesh is different segment, discuss uniqueness of
% paraperers
% Discuss regularization technices
When dealing with parameter estimation, you should always consider
questions about identifiability and uniqueness. This depends on the
model, the data and the objective functional
\cite{hadjicharalambous2015analysis}. A first test should be to
generate synthetic data with some prescribed parameters, feed this
error-free data to the data assimilation method, and ensure that you
are able to retrieve the parameters that generated that data. If you
end up with a different parameter set than the one generated the data,
your problem is not \emph{structurally identifiable}
\cite{chabiniok2016multiphysics}. Your problem is said to be
\emph{practical identifiable} if the parameters can be determined
uniquely based on the data at hand. If the test passes you should try
to add noise to the data, and see if your data assimilation is stable
with respect to noise in the measurement.

Demonstrating identifiability is 
often difficult, especially when your data is corrupted with noise,
and when the complexity of the model increases. Balancing the
complexity of the model in terms of \emph{model fidelity}, i.e whether
your 
model is rich enough to representing the data, and the identifiability
is often key in order to have robust model. A too simple model will
often fail to represent your data, while a too complex model, might
give completely different answers with just a slight perturbation of
the data, i.e with noise added to the measurement. The same problem is
encountered in other scientific disciplines. For example, in
statistical learning theory, \emph{overfitting} is related to
overparameterization of the model, in which you can perfectly
represent your data, but a small perturbation will cause a significant
error.

Another problem, which is also related to identifiability is
the question about uniqueness of the solution. A practical test is to
start the optimization algorithm from different initial points and see
if you end up at the same optimum. If this does not happen, it is
likely that the objective functional is not convex, which is a
requirement for uniqueness. Unfortunately, this is the case for many
problems in PDE-constrained optimization. In these cases, the objective
functional might have several distinct local minima, in which case
different techniques for finding the best local minima exists
\cite{farrell2015multiple}.

% \subsubsection{An example of a non-structurally identifiable problem}
% Split the ventricle in two, and use volume as the only input.
% Show that different parameter sets (with $\gamma$) will generate the
% same volume

% \subsubsection{Material parameter estimation}

% \paragraph{One volume measurement only}

% \paragraph{Multiple volume measurements}
